\LaTeX\ Author Guidelines for ICCV Proceedings

### 2021-03-10 chapters/supplementary.tex
Rewrite supplementary material and add proofs

\begin{algorithm*}
  \caption{Implementation of \algname{}, generalized algorithm for signed graph partitioning}
\hspace*{\algorithmicindent} \textbf{Input:} $\mathcal{G}(V,E,w^+,w^-)$ with $N$ nodes and $M$ edges; boolean \texttt{{\color{blue}addCannotLinkConstraints}} \\
\hspace*{\algorithmicindent} \textbf{Output:} Final clustering \\
  \hspace*{\algorithmicindent} 
  \begin{algorithmic}[1]
      \State $\tilde{\mathcal{G}}(\tilde{V},\tilde{E}) \gets \mathcal{G}(V,E,w^+,w^-)$  \Comment{Init. contracted graph}
      \State \texttt{UF} $\gets$ initUnionFind($V$) \Comment{Init. data structure representing clustering}
        \State PQ.push$(|w_e|, e) \quad \forall e \in E $  \Comment{Init. priority queue in desc. order of $|w_e|=|w_e^+ - w_e^-|$, $\mathcal{O}(|E|)$}
        \State \texttt{canBeMerged}$[e] \gets$ \texttt{True} $\,\,\, \forall e\in E$ \Comment{Init. cannot-link constraints}
    \State
      \While{PQ is \textbf{not} empty}
        \State $\tilde{w}, e_{uv} \gets $ PQ.popHighest() \Comment{$\mathcal{O}(\log |E|)$}
        \State \textbf{assert} \texttt{UF}.find($u$) $\neq$ \texttt{UF}.find($v$) \Comment{Edges in PQ always link nodes in different clusters}
        \If{({\color{green}\textbf{$\tilde{w} > 0$}}) \textbf{and} \texttt{canBeMerged}$[e_{uv}]$}
          \State PQ, \texttt{canBeMerged}, $\tilde{E}$ $\gets$ \textsc{UpdateNeighbors}($u,v$)
          \State $\tilde{V} \gets \tilde{V} \setminus \{ v\}$, $\quad \tilde{E} \gets \tilde{E} \setminus \{ e_{uv}\}$ \Comment{Update contracted graph}
          \State \texttt{UF}.merge($u,v$) \Comment{Merge clusters, $\mathcal{O}(\alpha(|E|))$}
        \ElsIf{({\color{red}\textbf{$\tilde{\cost} \leq 0$}}) \textbf{and} {\color{blue}\texttt{addCannotLinkConstraints}}}
          \State \texttt{canBeMerged}$[e_{uv}] \gets$ \texttt{False} \Comment{Constrain the two clusters}
        \EndIf
      \EndWhile
      \State
      \Return Final clustering given by union-find data structure  \texttt{UF}
  \end{algorithmic}
  \hspace*{1.5cm} 
    \begin{algorithmic}[1]
    \Function{UpdateNeighbors}{$u,v$}
      \State $\mathcal{N}_u = \{ t \in \tilde{V} | e_{ut}\in \tilde{E}  \}$
      \State $\mathcal{N}_v = \{ t \in \tilde{V} | e_{vt}\in \tilde{E}  \}$ 
      \For{$t \in \mathcal{N}_v$ } \Comment{Loop over neighbors in $\tilde{\mathcal{G}}$ of deleted node $v$}
        \State $\tilde{E} \gets \tilde{E} \setminus \{e_{vt}\}$
        \State $\tilde{w}_{vt} \gets$ PQ.delete($e_{vt}$) \Comment{$\mathcal{O}(\log |E|)$}
        \State \texttt{canBeMerged}$[e_{ut}] \gets$ \texttt{canBeMerged}$[e_{ut}]$ \textbf{and} \texttt{canBeMerged}$[e_{vt}]$
        \If{$t \in \mathcal{N}_u$ }\Comment{$t$ is a common neighbor of $u$ and $v$}
          \State $\tilde{w}_{ut} \gets$ PQ.delete($e_{ut}$)  \Comment{$\mathcal{O}(\log |E|)$}
          \State PQ.push($ |f(\tilde{w}_{ut}, \tilde{w}_{vt})|, e_{ut}$) \Comment{$\mathcal{O}(\log |E|)$  }
        \Else
          \State $\tilde{E} \gets \tilde{E} \cup \{e_{ut}\}$
          \State PQ.push($ |\tilde{w}_{vt}|, e_{ut}$) \Comment{$\mathcal{O}(\log |E|)$}
        \EndIf
      \EndFor
      \State
      \Return PQ, \texttt{canBeMerged}, $\tilde{E}$
    \EndFunction
  \end{algorithmic}
  \label{detailed_alg}
\end{algorithm*}

\section{Supplementary material}

\subsection{Implementation details and complexity of \algname{}} \label{sec:detailed_impl}

@@

\paragraph{Implementation} As we show in Algorithm \ref{detailed_alg}, our implementation of \algname{} is based on an union-find data structure and a heap allowing deletion of its elements. The algorithm starts with each node assigned to its own cluster and sorts all edges $e\in E$ in a heap/priority queue (PQ) by their absolute weight $|\cost_e|=|w_e^+ - w_e^-|$ in descending order, so that the most attractive and the most repulsive interactions are processed first. It then iteratively pops one edge $e_{uv}$ from PQ and, depending on the priority $\tilde{\cost}_{uv}$, does the following: in case of attractive interaction $\tilde{\cost}_{uv}>0$, provided that $e_{uv}$ was not flagged as a cannot-link constraint, then merge the connected clusters, perform an edge contraction of $e_{uv}$ in $\tilde{\mathcal{G}}_\Pi$ and update the priorities of new double edges as explained in Fig. \ref{fig:edge_contraction_and_contr_graph}. 
If, on the other hand, the interaction is repulsive ($\tilde{\cost}_{uv}\leq 0$) and the option \texttt{addCannotLinkContraints} of Alg. \ref{detailed_alg} is \texttt{True}, then the edge $e_{uv}$ is flagged as cannot-link constraint.

\paragraph*{Complexity} In the main loop, the algorithm iterates over all edges, but the only iterations presenting a complexity different from $\mathcal{O}(1)$ are the ones involving a merge of two clusters, which are at most $N-1$. By using a union-find data structure (with path compression and union by rank) the time complexity of \texttt{merge}$(u, v)$ and \texttt{find}($u$) operations is $\mathcal{O}(\alpha(N))$, where $\alpha$ is the slowly growing inverse Ackerman function. The algorithm then iterates over the neighbors of the merged cluster (at most $N$) and updates/deletes values in the priority queue ($\mathcal{O}(\log |E|)$). 
Therefore, similarly to a heap-based implementation of hierarchical agglomerative clustering, our implementation of \algname{} has a complexity of $\mathcal{O}(N^2 \log N)$. In the worst case, when the graph is dense and $|E|=N^2$, the algorithm requires $\mathcal{O}(N^2)$ memory. Nevertheless, in our practical applications the graph is much sparser, so $\mathcal{O}(|E|)=\mathcal{O}(N)$. 
With a single-linkage, corresponding to the choice of the \emph{Maximum} update rule in our framework, the algorithm can be clearly implemented by using the more efficient Kruskal's Minimum Spanning Tree algorithm with complexity $\mathcal{O}(N \log N)$. 
Moreover, in the next section, we present an efficient implementation of \algname{} with \emph{Absolute Maximum} linkage that has empirical $\mathcal{O}(N \log N)$ complexity. %\TODO{Empirical complexity plot?}
\begin{figure}[H]
        \centering
        \includegraphics[width=0.6\textwidth]{./figs/edge_contraction.pdf} % left bottom right top
    \centering
    \caption{Example of edge contraction. First row: original graph $\mathcal{G}$; clustering $\Pi$ (gray shaded areas) with dashed edges on cut; cannot-link constraints (violet bars). Second row: contracted graph $\tilde{\mathcal{G}}_\Pi$. In step ii), edge $e_{uv}$ is contracted and node $v$ deleted from $\tilde{\mathcal{G}}_\Pi$. In step iii), double edges $e_{tu}$ and $e_{tv}$ resulting from the edge contraction are replaced by a single edge with updated interaction.}\label{fig:edge_contraction_and_contr_graph}  
\end{figure}
\begin{table}[H]
\centering
    \small
    \vspace{-2em}
\begin{tabular}[t]{r | l }
            \toprule
            Linkage criteria & Update rule $f$ \\        
            \midrule
            Sum: & \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \tilde{\cost}_1+\tilde{\cost}_2$} \\ 
            \makecell[r]{Absolute \\Maximum:} & \thead[l]{
            $
            f(\tilde{\cost}_1,\tilde{\cost}_2) = \begin{cases} 
            \tilde{\cost}_1 & \text{if}\,\, |\tilde{\cost}_1|>|\tilde{\cost}_2|\\
            \tilde{\cost}_2 & \text{otherwise}
             \end{cases} 
            $}
               \\ 
            \makecell[r]{Average:} & \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \mathrm{weightAvg}\{ \tilde{\cost}_1, \tilde{\cost}_2 \} $}                 \\ 
            Maximum: & \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \max \{ \tilde{\cost}_1, \tilde{\cost}_2 \}  $} \\
            Minimum:& \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \min \{ \tilde{\cost}_1, \tilde{\cost}_2 \}  $} 
        \end{tabular}\vspace{1em}
        \caption{The table lists the update rules $f(\tilde{\cost}_1, \tilde{\cost}_2)$ associated to the linkage criteria of Table \ref{tab:linkage-criteria} and that are used to efficiently update the interactions between clusters.}
\label{tab:linkage_criteria_explicit}  
\end{table}

\begin{figure*}[t]
        \centering
        \includegraphics[width=0.7\textwidth]{./figs/proof_abs_max_example.pdf} % left bottom right top
\caption{\algname{} with \emph{AbsMax} linkage: Example representing the only case of edge contraction $e_{uv}$ that would introduce a positive attractive interaction between two constrained clusters. Note this can actually never happen with an \emph{AbsMax} linkage, because edge $e_{ut}$ has a lower absolute priority as compared to $e_{uv}$, so clusters $u$ and $t$ cannot have been constrained before $u$ and $v$ are merged.
}
\label{fig:abs_max_proof_example}  
\end{figure*}
\begin{algorithm*}[t]
  \caption{Mutex Watershed Algorithm proposed by \cite{wolf2018mutex}}
\hspace*{\algorithmicindent} \textbf{Input:} $\mathcal{G}(V,E,w^+,w^-)$ with $N$ nodes and $M$ edges \\
\hspace*{\algorithmicindent} \textbf{Output:} Final clustering \\
  \hspace*{\algorithmicindent} 
  \begin{algorithmic}[1]
      \State \texttt{UF} $\gets$ initUnionFind($V$) 
      \For{$(u,v)=e\in E$ in descending order of $|w_e|=|w_e^+ - w_e^-|$}
        \If{\texttt{UF}.find($u$) $\neq$ \texttt{UF}.find($v$)} \Comment{Check if $u,v$ are already in the same cluster}
          \If{({\color{green}\textbf{$w_e > 0$}}) \textbf{and} \texttt{canBeMerged}($u,v$)}  \Comment{Check for cannot-link constraints}
            \State \texttt{UF}.merge($u,v$) and inherit constraints of parent clusters
          \ElsIf{({\color{red}\textbf{$w_e \leq 0$}})}
            \State Add cannot-link constraints between parent clusters of $u,v$
          \EndIf
        \EndIf
      \EndFor
      \State
      \Return Final clustering given by union-find data structure \texttt{UF}
  \end{algorithmic}
  \label{alg:mutex_watershed}
\end{algorithm*}

\subsection{Properties of \algname{} with Absolute Maximum linkage}\label{sec:appendix_abs_max}
\paragraph{Remark on graph notation} The definition of a graph proposed by \cite{wolf2018mutex} makes a distinction between a set of positive edges $E^+$, associated with a set $W^+$ of positive scalar attributes representing merge affinities, and a set of negative edges $E^-$, associated with a set $W^-$ of positive attributes representing split tendencies. On the other hand, in our definition $\mathcal{G}(V,E,w^+,w^-)$ each edge have both an attractive $w_e^+$ and a repulsive $w_e^-$ attribute, so we can make them equivalent by defining:
\begin{align}
E^+ =&\,\, \{ e \in E \,\,\text{s.t.} \,\,w_e = w_e^+ - w_e^- > 0\}, \\
E^- =&\,\, \{ e \in E \,\,\text{s.t.}\,\, w_e = w_e^+ - w_e^- \leq 0\}, \\
W^+ =&\,\, \{ |w_e| \,\,\text{s.t.}\,\, e \in E^+\},\\
 W^- =&\,\, \{ |w_e| \,\,\text{s.t.}\,\, e \in E^-\}.
\end{align}

\begin{prop} \label{prop:equiv_MWS}
The Mutex Watershed Algorithm \ref{alg:mutex_watershed} (MWS) with empirical \linebreak $\mathcal{O}(N \log N)$ complexity introduced by \cite{wolf2018mutex} returns the same final clustering given by the \algname{} Algorithm \ref{detailed_alg} with the use of cannot-link constraints and an Absolute Maximum update rule:
\begin{equation}\label{eq:def_abs_max}
f_{\mathrm{Abs.Max.}}(\tilde{\cost}_1,\tilde{\cost}_2) = \begin{cases} 
            \tilde{\cost}_1 & \text{if}\,\, |\tilde{\cost}_1|>|\tilde{\cost}_2|\\
            \tilde{\cost}_2 & \text{otherwise}
             \end{cases} 
\end{equation}
\end{prop}
\begin{proof}
Both algorithms sort edges in descending order of the absolute interactions $|w_e|$ and then iterate over all of them. The only difference is that MWS, after merging two clusters, does not update the interactions between the new cluster and its neighbors. 
However, since with an Abs. Max. linkage the interaction between clusters is simply given by the edge with highest absolute weight $|w_e|$, the order by which edges are iterated over in \algname{} is never updated. Thus, both algorithms perform precisely the same steps and return the same clustering.
\end{proof}
\begin{prop} \label{prop:abs_max_cannot_link_property}
The \algname{} Algorithm \ref{detailed_alg} with the Absolute Maximum linkage defined in Eq. \ref{eq:def_abs_max} returns the same final clustering whether or not cannot-link constraints are enforced. 
\end{prop}
\begin{proof}
In the \algname{} Algorithm \ref{detailed_alg}, the clustering is updated only when two clusters are merged and the condition at line 9 is satisfied. 
We also observe that, in the unconstrained version of \algname{}, the predicate \texttt{canBeMerged} at line 9 can never be false because cannot-link constraints are never introduced at line 14.
Let us now contradict the initial hypothesis and assume by absurd that the constrained version of \algname{} introduces a cannot-link constraints between two clusters sharing a positive interaction $\tilde{w}>0$ and outputs a different clustering as compared to the unconstrained version. 
This can happen only in the situation shown in Fig. \ref{fig:abs_max_proof_example}, when two clusters $u$ and $v$ are merged together and share a common neighboring node $t$ having the following two properties: a) $u$ and $t$ are already constrained and share a repulsive interaction $w_{ut}\leq0$, b) $v$ and $t$ share an attractive interaction $w_{vt}>0$ that is higher in absolute value $|w_{vt}|>|w_{ut}|$. 
Then, according to Eq. \ref{eq:def_abs_max}, the new merged cluster $uv$ and $t$ are constrained and share a positive interaction. 
But this case can never happen, since if $|w_{vt}|>|w_{ut}|$ then clusters $v$ and $t$ are merged before clusters $u$ and $t$ are constrained.  
\end{proof}

\subsection{Predicting signed edge weights with a CNN}
Our CNN model outputs affinities in the form of pseudo-probabilities $p:E \rightarrow [0,1]$, where $p=0$ represents a boundary evidence. In order to use them as input of the algorithms in our framework, we mapped them to positive and negative values\footnote{Note that in general attractive and repulsive interactions $w^+$ and $w^-$ can be independently estimated with different classifiers.}. The most common approaches use \emph{additive} \cite{ailon2008aggregating} or \emph{logarithmic} \cite{finkel2008enforcing,andres2012globally} mappings:
\begin{align} 
\cost_{e,\mathrm{Add}} =&\,\, p_e - \beta, \\
\cost_{e,\mathrm{Log}} =&\,\, \log \left( \frac{p_e}{1-p_e} \right) - \log \left( \frac{\beta}{1-\beta} \right), \label{eq:mappings}
\end{align}
where $\beta \in [0,1]$ is a \emph{bias} parameter that allow a tuning between over- and under-segmentation. We evaluated both of them empirically with each of the tested linkage and found that the additive mapping is the best option in all cases apart from the \emph{Sum} linkage. Note that varying the parameter $\beta$ does not usually define a hierarchy of nested clusterings, thus it is not equivalent to varying a threshold parameter in HAC. This hierarchical property is only valid for \algname{} without constraints and with \emph{Average}, \emph{Max} or \emph{Min} linkage.

\subsection{Neuron segmentation and compared methods}\label{sec:cremi_details}
\paragraph{Training details} The data from the CREMI challenge is highly anisotropic and contains artifacts like missing sections, staining precipitations and support film folds. 
To alleviate difficulties stemming from misalignment, we use a version of the data that was elastically realigned by the challenge organizers with the method of \cite{saalfeld2012elastic}.
We train a 3D U-Net \cite{ronneberger2015u,cciccek20163d} using the same architecture as \cite{funke2018large} and predict long-and-short range affinities 
as described in \cite{lee2017superhuman}. In addition to the standard data augmentation techniques of random rotations, random flips and  elastic deformations, we simulate data artifacts.
In more detail, we randomly zero-out slices, decrease the contrast of slices, simulate tears, introduce alignment jitter and paste artifacts extracted from the training data. Both \cite{funke2018large} and \cite{lee2017superhuman} have shown
that these kinds of augmentations can help to alleviate issues caused by EM-imaging artifacts.
We use L2 loss and Adam optimizer to train the network. The model was trained on all the three samples with available ground truth labels.

\paragraph{THRESH and WSDT} The basic post-processing methods we consider cannot take long-range affinities into account, so we only consider direct neighbors affinities and generate a boundary map by taking an average over the 3 directions. Based on this boundary map, we run connected components (THRESH) and we also run a watershed algorithm  seeded at the maxima of the smoothed distance transform (WSDT). For WSDT, the degree of smoothing was optimized such that each region receives as few seeds as possible, without however causing severe under-segmentation. Due to the anisotropy of the data, we generate 2D WSDT superpixels by considering each 2D image in the stack singularly.

\paragraph{Multi-step pipelines} Given the 2D WSDT superpixels, we build a 3D region-adjacency graph such that each node represents a superpixel. The weights of the edges connecting neighboring superpixels are computed by taking an average over both short- and long-range affinities connecting the two regions. We then convert the edge probabilities to signed weights using the logarithmic mapping defined in Eq. \ref{eq:mappings} and solve the multicut problem on this graph. For our experiments, we use the approximate Kernighan-Lin solver \cite{keuper2015efficient,kernighan1970efficient} (WSDT+MC). In some cases, the long-range affinities predicted by the CNN can connect two superpixels that are not direct-neighbors. Thus, in these cases we introduce additional \emph{lifted} edges in the graph and an instance of the lifted multicut problem (WSDT+LMC). This time, similarly to the methods mentioned in \cite{beier2016efficient}, we used a combination of approximate solvers consisting in GAEC and Kernighan-Lin.

>>>>

\renewcommand{\thesection}{A\arabic{section}}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

\section{Appendix}

\subsection{Implementation and complexity of \algname{}} \label{sec:detailed_impl}

@@

\afterpage{ %
  \begin{algorithm*}[p]
    \caption{Implementation of \algname{} - Phase 1}
  \hspace*{\algorithmicindent} \textbf{Input:} $\mathcal{G}(V,E,w^+,w^-)$ with $N$ nodes and $M$ edges; boolean \texttt{{\color{blue}addCannotLinkConstraints}} \\
  \hspace*{\algorithmicindent} \textbf{Output:} Final clustering \\
    \hspace*{\algorithmicindent} 
    \begin{algorithmic}[1]
        \State $\tilde{\mathcal{G}}(\tilde{V},\tilde{E}) \gets \mathcal{G}(V,E,w^+,w^-)$  \Comment{Init. contracted graph}
        \State \texttt{UF} $\gets$ initUnionFind($V$) \Comment{Init. data structure representing clustering}
          \State PQ.push$(|w_e|, e) \quad \forall e \in E $  \Comment{Init. priority queue in desc. order of $|w_e|=|w_e^+ - w_e^-|$, $\mathcal{O}(|E|)$}
          \State \texttt{canBeMerged}$[e] \gets$ \texttt{True} $\,\,\, \forall e\in E$ \Comment{Init. cannot-link constraints}
      \State
        \While{PQ is \textbf{not} empty}
          \State $\tilde{w}, e_{uv} \gets $ PQ.popHighest() \Comment{$\mathcal{O}(\log |E|)$}
          \State \textbf{assert} \texttt{UF}.find($u$) $\neq$ \texttt{UF}.find($v$) \Comment{Edges in PQ always link nodes in different clusters}
          \If{({\color{green}\textbf{$\tilde{w} > 0$}}) \textbf{and} \texttt{canBeMerged}$[e_{uv}]$}
            \State PQ, \texttt{canBeMerged}, $\tilde{E}$ $\gets$ \textsc{UpdateNeighbors}($u,v$)
            \State $\tilde{V} \gets \tilde{V} \setminus \{ v\}$, $\quad \tilde{E} \gets \tilde{E} \setminus \{ e_{uv}\}$ \Comment{Update contracted graph}
            \State \texttt{UF}.merge($u,v$) \Comment{Merge clusters, $\mathcal{O}(\alpha(|E|))$}
          \ElsIf{({\color{red}\textbf{$\tilde{\cost} \leq 0$}}) \textbf{and} {\color{blue}\texttt{addCannotLinkConstraints}}}
            \State \texttt{canBeMerged}$[e_{uv}] \gets$ \texttt{False} \Comment{Constrain the two clusters}
          \EndIf
        \EndWhile
        \State
        \Return Final clustering given by union-find data structure  \texttt{UF}
    \end{algorithmic}
    \hspace*{1.5cm} 
      \begin{algorithmic}[1]
      \Function{UpdateNeighbors}{$u,v$}
        \State $\mathcal{N}_u = \{ t \in \tilde{V} | e_{ut}\in \tilde{E}  \}$
        \State $\mathcal{N}_v = \{ t \in \tilde{V} | e_{vt}\in \tilde{E}  \}$ 
        \For{$t \in \mathcal{N}_v$ } \Comment{Loop over neighbors in $\tilde{\mathcal{G}}$ of deleted node $v$}
          \State $\tilde{E} \gets \tilde{E} \setminus \{e_{vt}\}$
          \State $\tilde{w}_{vt} \gets$ PQ.delete($e_{vt}$) \Comment{$\mathcal{O}(\log |E|)$}
          \State \texttt{canBeMerged}$[e_{ut}] \gets$ \texttt{canBeMerged}$[e_{ut}]$ \textbf{and} \texttt{canBeMerged}$[e_{vt}]$
          \If{$t \in \mathcal{N}_u$ }\Comment{$t$ is a common neighbor of $u$ and $v$}
            \State $\tilde{w}_{ut} \gets$ PQ.delete($e_{ut}$)  \Comment{$\mathcal{O}(\log |E|)$}
            \State PQ.push($ |f(\tilde{w}_{ut}, \tilde{w}_{vt})|, e_{ut}$) \Comment{$\mathcal{O}(\log |E|)$  }
          \Else
            \State $\tilde{E} \gets \tilde{E} \cup \{e_{ut}\}$
            \State PQ.push($ |\tilde{w}_{vt}|, e_{ut}$) \Comment{$\mathcal{O}(\log |E|)$}
          \EndIf
        \EndFor
        \State
        \Return PQ, \texttt{canBeMerged}, $\tilde{E}$
      \EndFunction
    \end{algorithmic}
    \label{detailed_alg}
  \end{algorithm*}
  % \begin{figure*}[t]
  %         \centering
  %         \includegraphics[width=0.7\textwidth]{./figs/proof_abs_max_example.pdf} % left bottom right top
  % \caption{\algname{} with \emph{AbsMax} linkage: Example representing the only case of edge contraction $e_{uv}$ that would introduce a positive attractive interaction between two constrained clusters. Note this can actually never happen with an \emph{AbsMax} linkage, because edge $e_{ut}$ has a lower absolute priority as compared to $e_{uv}$, so clusters $u$ and $t$ cannot have been constrained before $u$ and $v$ are merged. \TODO{Remove?}
  % }
  % \label{fig:abs_max_proof_example}  
  % \end{figure*}
  \begin{algorithm*}[p]
    \caption{Mutex Watershed Algorithm proposed by \cite{wolf2018mutex}}
  \hspace*{\algorithmicindent} \textbf{Input:} $\mathcal{G}(V,E,w^+,w^-)$ with $N$ nodes and $M$ edges \\
  \hspace*{\algorithmicindent} \textbf{Output:} Final clustering \\
    \hspace*{\algorithmicindent} 
    \begin{algorithmic}[1]
        \State \texttt{UF} $\gets$ initUnionFind($V$) 
        \For{$(u,v)=e\in E$ in descending order of $|w_e|=|w_e^+ - w_e^-|$}
          \If{\texttt{UF}.find($u$) $\neq$ \texttt{UF}.find($v$)} \Comment{Check if $u,v$ are already in the same cluster}
            \If{({\color{green}\textbf{$w_e > 0$}}) \textbf{and} \texttt{canBeMerged}($u,v$)}  \Comment{Check for cannot-link constraints}
              \State \texttt{UF}.merge($u,v$) and inherit constraints of parent clusters
            \ElsIf{({\color{red}\textbf{$w_e \leq 0$}})}
              \State Add cannot-link constraints between parent clusters of $u,v$
            \EndIf
          \EndIf
        \EndFor
        \State
        \Return Final clustering given by union-find data structure \texttt{UF}
    \end{algorithmic}
    \label{alg:mutex_watershed}
  \end{algorithm*}
  \clearpage
}

\paragraph{Implementation} Our implementation of \algname{} is based on an union-find data structure and a heap allowing deletion of its elements. 
In Phases 2 and 3, \algname{} is equivalent to a standard hierarchical agglomerative clustering algorithm for unsigned graphs. In Algorithm \ref{detailed_alg}, we show our implementation of Phase 1, involving cannot-link constraints.
In phase 1, the algorithm starts with each node assigned to its own cluster and sorts all edges $e\in E$ in a heap/priority queue (PQ) by their absolute weight $|\cost_e|=|w_e^+ - w_e^-|$ in descending order, so that the most attractive and the most repulsive interactions are processed first. It then iteratively pops one edge $e_{uv}$ from PQ and, depending on the priority $\tilde{\cost}_{uv}$, does the following: in case of attractive interaction $\tilde{\cost}_{uv}>0$, provided that $e_{uv}$ was not flagged as a cannot-link constraint, then merge the connected clusters, perform an edge contraction of $e_{uv}$ in $\tilde{\mathcal{G}}_\Pi$ and update the priorities of new double edges as explained in Fig. \ref{fig:edge_contraction_and_contr_graph}. 
If, on the other hand, the interaction is repulsive ($\tilde{\cost}_{uv}\leq 0$) and the option \texttt{addCannotLinkContraints} of Alg. \ref{detailed_alg} is \texttt{True}, then the edge $e_{uv}$ is flagged as cannot-link constraint.

\begin{table}[t]
\centering
    \small
    % \vspace{-2em}
\begin{tabular}[t]{r | l }
            \toprule
            Linkage criteria & Update rule $f$ \\        
            \midrule
            Sum: & \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \tilde{\cost}_1+\tilde{\cost}_2$} \\ 
            \makecell[r]{Absolute \\Maximum:} & \thead[l]{
            $
            f(\tilde{\cost}_1,\tilde{\cost}_2) = \begin{cases} 
            \tilde{\cost}_1 & \text{if}\,\, |\tilde{\cost}_1|>|\tilde{\cost}_2|\\
            \tilde{\cost}_2 & \text{otherwise}
             \end{cases} 
            $}
               \\ 
            \makecell[r]{Average:} & \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \mathrm{weightAvg}\{ \tilde{\cost}_1, \tilde{\cost}_2 \} $}                 \\ 
            Single: & \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \max \{ \tilde{\cost}_1, \tilde{\cost}_2 \}  $} \\
            Complete:& \thead[l]{$f(\tilde{\cost}_1,\tilde{\cost}_2) = \min \{ \tilde{\cost}_1, \tilde{\cost}_2 \}  $} 
        \end{tabular}\vspace{1em}
        \caption{The table lists the update rules $f(\tilde{\cost}_1, \tilde{\cost}_2)$ associated to the linkage criteria of Table \ref{tab:linkage-criteria} and that are used to efficiently update the interactions between clusters.}
\label{tab:linkage_criteria_explicit}  
\end{table}
\begin{figure}[t]
        \centering
        \includegraphics[width=0.48\textwidth]{./figs/edge_contraction.pdf} % left bottom right top
    \centering
    \caption{Example of edge contraction. First row: original graph $\mathcal{G}$; clustering $\Pi$ (gray shaded areas) with dashed edges on cut; cannot-link constraints (violet bars). Second row: contracted graph $\tilde{\mathcal{G}}_\Pi$. In step ii), edge $e_{uv}$ is contracted and node $v$ deleted from $\tilde{\mathcal{G}}_\Pi$. In step iii), double edges $e_{tu}$ and $e_{tv}$ resulting from the edge contraction are replaced by a single edge with updated interaction.}\label{fig:edge_contraction_and_contr_graph}  
\end{figure}

\paragraph*{Complexity} 
In the main loop of Phase 1, the algorithm iterates over all edges, but the only iterations presenting a complexity different from $\mathcal{O}(1)$ are the ones involving a merge of two clusters, which are at most $N-1$. By using a union-find data structure (with path compression and union by rank) the time complexity of \texttt{merge}$(u, v)$ and \texttt{find}($u$) operations is $\mathcal{O}(\alpha(N))$, where $\alpha$ is the slowly growing inverse Ackerman function. The algorithm then iterates over the neighbors of the merged cluster (at most $N$) and updates/deletes values in the priority queue ($\mathcal{O}(\log |E|)$). 
Therefore, 
similarly to a heap-based implementation of hierarchical agglomerative clustering, our implementation of \algname{} has a complexity of $\mathcal{O}(N^2 \log N)$. In the worst case, when the graph is dense and $|E|=N^2$, the algorithm requires $\mathcal{O}(N^2)$ memory. Nevertheless, in our practical applications the graph is much sparser, so $\mathcal{O}(|E|)=\mathcal{O}(N)$. 
With a single-linkage, corresponding to the choice of the \emph{Maximum} update rule in our framework, the algorithm can be clearly implemented by using the more efficient Kruskal's Minimum Spanning Tree algorithm with complexity $\mathcal{O}(N \log N)$. 
Moreover, \algname{} with \emph{Absolute Maximum} linkage can be implemented more efficiently (see next section). %\TODO{Empirical complexity plot?}

\subsection{Proofs of Propositions \ref{prop:absmax_mutex}, \ref{prop:weight_shift_invariant}, \ref{prop:ultraMetric1}, and \ref{prop:ultraMetric2}  }
\label{sec:proposition_proofs}
% \subsubsection{Mutex Watershed and AbsMax linkage}\label{sec:appendix_abs_max}
\begin{lemma} \label{lemma:absMax_and_complete_property}
Let us consider the \algname{} Algorithm \ref{main_alg} with AbsMax or Complete linkage criteria (see linkage definition in Table \ref{tab:linkage-criteria}). If the algorithm enforces a constraint between two clusters in \emph{Phase 1}, then the interaction between the clusters will never become positive over the course of the following agglomeration steps. Moreover, we also have that with any of these two linkages: i) \algname{} returns the same final clustering whether or not cannot-link constraints are enforced; ii) no clusters are merged in \emph{Phase 2} (after constraints have been removed). 
\end{lemma}
\begin{proof}
First, we note that two clusters are constrained in \emph{Phase 1} only if their interaction is repulsive. We then also observe that with complete linkage (see Table~\ref{tab:linkage_criteria_explicit}) the signed interaction between two clusters can only decrease over the course of the agglomeration. Thus, if two clusters are constrained by the algorithm, their negative interaction cannot increase and become positive later on in the agglomeration process.

If, on the other hand, we consider \algname{} with a Absolute Maximum linkage (see Table~\ref{tab:linkage_criteria_explicit}), during the agglomeration the interaction between two clusters can only increase in absolute value. Thus, the negative interaction $\interact{}(S_i,S_j)<0$ between two constrained clusters can possibly become positive over the course of next agglomeration steps only if there is at least another pair of clusters with positive interaction $\interact{}(S_l,S_t)>0$ higher in absolute value: $|\interact{}(S_l,S_t)| > |\interact{}(S_i,S_j)|$.
If such clusters $S_l, S_t$ with positive interaction exist, we note that they must also be constrained (in the opposite case, the algorithm would have already merged them before to constrain $S_i$ and $S_j$, because their priority was higher). In other words, a constrained negative interaction can become positive only if there is already another positive constrained interaction: but this can never be the case because initially all constrained interactions are negative.

Finally, we prove that \algname{} with either AbsMax or Complete linkage returns the same final clustering whether or not cannot-link constraints are enforced. In Algorithm \ref{detailed_alg}, the interactions between clusters are updated only when two clusters are merged and the condition at line 9 is satisfied (i.e. when an interaction is both positive and not constrained). However, we have just shown that with these two linkage methods an interaction can never be both positive and constrained at the same time, so we directly conclude that the constrained and unconstrained versions of the algorithm will perform precisely the same agglomeration steps.
\end{proof}

% We now prove Proposition \ref{prop:absmax_mutex}:

\absmaxmutex*

% \begin{prop} \label{prop:equiv_MWS}
% The Mutex Watershed Algorithm \ref{alg:mutex_watershed} (MWS) with empirical \linebreak $\mathcal{O}(N \log N)$ complexity introduced by \cite{wolf2018mutex} returns the same final clustering given by the \algname{} Algorithm \ref{detailed_alg} with the use of cannot-link constraints and an Absolute Maximum update rule:
% \begin{equation}\label{eq:def_abs_max}
% f_{\mathrm{Abs.Max.}}(\tilde{\cost}_1,\tilde{\cost}_2) = \begin{cases} 
%             \tilde{\cost}_1 & \text{if}\,\, |\tilde{\cost}_1|>|\tilde{\cost}_2|\\
%             \tilde{\cost}_2 & \text{otherwise}
%              \end{cases} 
% \end{equation}
% \end{prop}

% \paragraph{Remark on graph notation in \cite{wolf2018mutex}} The definition of a graph proposed by \cite{wolf2018mutex} makes a distinction between a set of positive edges $E^+$, associated with a set $W^+$ of positive scalar attributes representing merge affinities, and a set of negative edges $E^-$, associated with a set $W^-$ of positive attributes representing split tendencies. On the other hand, in our definition $\mathcal{G}(V,E,w^+,w^-)$ each edge have both an attractive $w_e^+$ and a repulsive $w_e^-$ attribute, so we can make them equivalent by defining:
% \begin{align}
% E^+ =&\,\, \{ e \in E \,\,\text{s.t.} \,\,w_e = w_e^+ - w_e^- > 0\}, \\
% E^- =&\,\, \{ e \in E \,\,\text{s.t.}\,\, w_e = w_e^+ - w_e^- \leq 0\}, \\
% W^+ =&\,\, \{ |w_e| \,\,\text{s.t.}\,\, e \in E^+\},\\
%  W^- =&\,\, \{ |w_e| \,\,\text{s.t.}\,\, e \in E^-\}.
% \end{align}
